Family Fortunes: A Quantitative Analysis of the Early Iron Age Cemeteries at Knossos, Crete

### 2019-04-01 analysis/paper/paper.Rmd
Writing methods and analysis

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  fig.retina = 2,
  comment = "#>",
  fig.path = "../figures/"
)
```

@@

A first important step in preparing the data for analysis was the estimation of the number of individuals buried in each tomb. The method chosen for this estimation broadly follows that of Cavanagh [-@Cavanagh1996, 659-660], who took the number of pithoi (the most common urn type) and the number of pithoi, amphorae and kraters (which were in some cases used as urns) as middle and upper estimates of the number of interments, with all securely identified interments (in the form of ashes or osteological material) being the lower limit. In the present study, the mean of these three estimates was taken, to provide a single aggregate measure for the purposes of analysis (where no interments were known, this variable was still assigned a value of "1", for all tombs must have contained at least one burial).

@@

library(ineq)
library(ggrepel)
library(reshape)
library(data.table)
library(extrafont)
library(ggplot2)
library(RColorBrewer)
library(grid)
library(dplyr)
font_import(paths="../paper/font",pattern="GARA")
loadfonts(device="win")

@@

For analysis at the level of individual finds, however, pithoi alone were taken as the best proxy, as they correlated most closely with the mean estimate, producing a pearson correlation coefficient of `r cor_burial_proxy` (Figure \@ref(fig:burialestimate)). For the bulk of the analyses, any tombs lacking identified remains or a single vessel suitable for use as an urn were not considered, thus excluding the most heavily damaged and looted tombs.

```{r aorist_all_cems}

@@

ggplot(aorist_all_tombs,aes(decade,aoristic_density))+
  geom_col()+xlim(1100,630)+theme_minimal()+
        theme(legend.title=element_blank(),
          text=element_text(family="Garamond", size=12))+
              labs(x="Date (Years BC)", y="Aoristic Density")+
                 scale_fill_brewer(palette = "Set2")

@@

```{r aorist_uncertainty, fig.dim = c(7, 4)}

pots<-read.csv("../data/raw_data/KnossosPots.csv",header=T)
pots<-subset(pots,pots$cemetery=="FOR"|pots$cemetery=="TK"|pots$cemetery=="NC")

View(aorist_pot_medians)

ggplot(aorist_melt,aes(variable,value, fill=period))+geom_boxplot(varwidth=TRUE)+theme_bw()+
      theme(legend.title=element_blank(),
        text=element_text(family="Garamond", size=12),
        axis.text.x = element_text(angle = 70, hjust = 1),
        panel.grid.major.x = element_blank())+
            labs(x="Date (Years BC)", y="Aoristic Sums per Tomb")+
               scale_fill_brewer(palette = "Set3")+
                scale_y_continuous(expand = c(0, 0),limits=c(0,20))

tombs_gini<-matrix(0, ncol = 0, nrow = 1)
tombs_gini<-data.frame(tombs_gini)
for(i in colnames(aorist_tomb_values[2:51])) {
  tombs_gini[[paste0(i)]] <- ineq(aorist_tomb_values[[i]],type="Gini")
}

ggplot() + 
  geom_rect(data = rects, aes(xmin = xstart, xmax = xend, ymin = -Inf, ymax = Inf, fill = col), alpha = 0.4) +
     geom_line(data = tombs_gini, aes(decade,gini_coefficient))+
      theme_bw()+theme(legend.title=element_blank(),
          text=element_text(family="Garamond", size=12))+
              labs(x="Date (Years BC)", y="Gini Coefficient")+
                 scale_fill_brewer(palette = "Set3")+ coord_cartesian(
                    xlim = c(1100, 600), ylim = c(0.4,0.8))+ scale_x_reverse(expand = c(0, 0),limits=c(1100,600))

@@

@@

@@

pithoi<-subset(pots,pots$shape=="Pithos")
all_urns<-subset(pots,pots$shape=="Pithos"|pots$shape=="Amphora"|pots$shape=="Krater")

>>>>

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  fig.retina = 2,
  comment = "#>",
  fig.path = "../figures/",
  dev="png"
)
```

@@

A first important step in preparing the data for analysis was the estimation of the number of individuals buried in each tomb. The method chosen for this estimation broadly follows that of Cavanagh [-@Cavanagh1996, 659-660], who took the number of pithoi (the most common urn type) and the number of pithoi, amphorae and kraters (which were in some cases used as urns) as middle and upper estimates of the number of interments, with all securely identified interments (in the form of ashes or osteological material) being the lower limit (Figure \@ref(fig:burialestimate)). In the present study, the mean of these three estimates was taken, to provide a single aggregate measure for the purposes of analysis (where no interments were known, this variable was still assigned a value of "1", for all tombs must have contained at least one burial).

@@

library(ggpubr)
library(cowplot)
library(gridExtra)
library(ineq)
library(ggrepel)
library(reshape)
library(data.table)
library(extrafont)
library(ggplot2)
library(RColorBrewer)
library(grid)
library(dplyr)
font_import(paths="../paper/font",pattern="GARA")
loadfonts(device="win")

@@

For most diachronic analyses, however, pithoi alone were taken as the best proxy for burial numbers, as they correlated most closely with the mean estimate, (Pearson's *r* = `r cor_burial_proxy`). In most cases, and often simply by virute of the methods employed, any tombs lacking identified remains or a single vessel suitable for use as an urn were not considered, thus excluding what are presumably among the most heavily damaged and looted tombs.

##Aoristic approaches to dating

A central concern of the present study was how to deal with the temporal uncertainty implicit within the assigning of individual vessels to stylistically-defined periods. A vessel dated to the Protogeometric B period could, on our best estimates of absolute chronology, date to any point between 840 and 810 BC. This situation results in major challenges to the visualisation and analysis of such data through mean which normally rely on event- and not range-based data inputs. A simple, if unsatisfying, solution is to use the midpoint of the range asigned to each vessel (i.e. 825 for our PGB example). This method is used in a couple of instances in the present paper when the course-grained sorting of the data (into 50 year periods) renders the innacuracies of such a method less significant.

```{r allpithoi, fig.cap="Counts of pithoi across both cemeteries through time. Vessel dates have here been based on the midpoint of the absolute date range accompanying their respective ceramic periods."}

pots<-read.csv("../data/raw_data/KnossosPots.csv",header=T)
pots<-subset(pots,pots$cemetery=="FOR"|pots$cemetery=="TK"|pots$cemetery=="NC")
pithoi<-subset(pots,pots$shape=="Pithos")
all_urns<-subset(pots,pots$shape=="Pithos"|pots$shape=="Amphora"|pots$shape=="Krater")

ggplot(pots, aes(ave_date,fill="Pots"))+geom_histogram()+
  theme_bw()+theme(legend.title=element_blank(),
      text=element_text(family="Garamond", size=12))+
          labs(x="Date (Years BC)", y="Total Vessel Counts")+
              scale_fill_brewer(palette = "Set3")+ 
                theme(legend.position="none")+
                  scale_y_continuous(expand=c(0,0),limits=c(0,600))+
                    scale_x_reverse(expand = c(0, 0),limits=c(1100,600))

But the artificial clustering this technique engenders around select absolute dates proves misleading when, for instance, the total number of pithoi deposited by date are plotted via a histogram (Figure \@ref(fig:allpithoi)); the apparent cessation of burial activity for 30 years in the 8th century, followed by the most intense episode of deposition yet seen, is simply a fiction generated by the processing of the data.

A alternative way of dealing with temporal uncertainty, and the method employed extensively in this paper, is the use of so-called "aoristic analysis". Here, rather than assign a pot a single date, the entire timespan under investigation is divided into equal chunks (in this case, decades), and probability values summing to 1 are then assigned for each vessel, per chunk of time, based on its accordant date range. So, our PGB vessel, dateable between 840 and 810 BC would register a value of 0.33 for each of the three decades within which we it could theoretically have been produced. These values for individual vessels can can then simply be summed to produce tomb- or cemetery-level signatures. The greatest contribution of this method is that it incorporates temporal uncertainty into the size of the resulting values, that is, "events with tight temporal definition contribute more to the total probability over their range than do loosely defined events" (JOHNSON 2003 p2).

```{r aoristallcems, fig.cap="Aoristic sums of all vessels across both cemeteries. This represents a probabilistically weighted visualisation of depositional activity through time."}

@@

ggplot(aorist_all_tombs,aes(decade,aoristic_density, fill="Aoristic Sum"))+
  geom_col()+theme_bw()+theme(legend.title=element_blank(),
      text=element_text(family="Garamond", size=12))+
          labs(x="Date (Years BC)", y="Aoristic Sum")+
              scale_fill_brewer(palette = "Set3")+ 
                theme(legend.position="none")+
                  scale_y_continuous(expand=c(0,0),limits=c(0,225))+
                    scale_x_reverse(expand = c(0, 0),limits=c(1100,600))

@@

As seen in Figure \@ref(fig:aoristallcems), the artifical peaks and troughs of the simple averaging method are avoided, and different patterns become visible. Though these methods present their own limitations, which will be explored in the following sections, they are considered a more robust and productive approach to the temporal uncertainty of ceramic dating and, as such, are used for much of the present analysis.

##Theory and method

Before proceeding to the main discussion, it should be stressed that, though I hope to demonstrate the power of quantitative and statistical approaches to the mortuary record, data-points cannot replace description, and theory must not be absent in the use of such methods. Quantifications and statistical analyses should, ideally, be employed as tools for the development and exploration of theoretically-informed models of social practice. Throughout the following analysis, I aim to link the observed patterning in the material back to possible forms of human behaviour, which in turn guides the focus of subsequent analyses, and so on. It is hoped that the value of such an approach will become clear.

#Analysis

##Introduction
The great advantage of the large, quantitative dataset compiled for the present analysis is that it permits investigation of several nested levels of potentially significant patterning. From individual burials, to collective tombs, to the cemeteries as a whole, and between cemeteries in the landscape, we have the opportunity to study and compare both the chronological and contemporary variation across these various levels. The following analysis is divided broadly into these two distinct, though complementary, perspectives. I consider, firstly, how temporal variations in the material may relate to broader social dynamics and, secondly, how the form and content of the tombs and their assemblages can enrich the schematic picture derived from the chronological patterns. Taken together, these two approaches enable us to develop a nuanced account of the changing social landscape of EIA Knossos, drawing together issues of group identity, external relations, and the communication of wealth and status through funerary practice.

##Temporal variation in the Knossian Cemeteries

An obvious place to begin is with the chronology of the burials themselves. The rate and number of burials being made through time provides the background against which to consider all other developments.

```{r aoristbycem, fig.cap="Aorist sums of pithoi (left) as a proxy for cremation burials, and all vessels (right). The sums are colour coded to show the values of the two cemeteries, and the total values.", fig.dim=c(8,4)}
KNC<-subset(tombs,tombs$cem_condensed=="NC")

aorist_KNC_pithoi <- colSums(KNC[207:256])
aorist_KNC_pithoi<-data.frame(aorist_KNC_pithoi)

colnames(aorist_KNC_pithoi)<-c("aoristic_density")
row.names(aorist_KNC_pithoi)<-dates
aorist_KNC_pithoi<-data.frame(aorist_KNC_pithoi)
setDT(aorist_KNC_pithoi, keep.rownames = "decade")
aorist_KNC_pithoi$decade<-as.numeric(aorist_KNC_pithoi$decade)

aorist_FOR_pithoi <-colSums(FOR[207:256])
aorist_FOR_pithoi<-data.frame(aorist_FOR_pithoi)

colnames(aorist_FOR_pithoi)<-c("aoristic_density")
row.names(aorist_FOR_pithoi)<-dates
aorist_FOR_pithoi<-data.frame(aorist_FOR_pithoi)
setDT(aorist_FOR_pithoi, keep.rownames = "decade")
aorist_FOR_pithoi$decade<-as.numeric(aorist_FOR_pithoi$decade)

aorist_all_pithoi <- colSums(tombs[207:256])
aorist_all_pithoi<-data.frame(aorist_all_pithoi)

colnames(aorist_all_pithoi)<-c("aoristic_density")

row.names(aorist_all_pithoi)<-dates
aorist_all_pithoi<-data.frame(aorist_all_pithoi)
setDT(aorist_all_pithoi, keep.rownames = "decade")
aorist_all_pithoi$decade<-as.numeric(aorist_all_pithoi$decade)

aorist_pithoi_plot<-ggplot(NULL)+
  geom_col(data=aorist_all_pithoi,aes(decade,aoristic_density,fill=" All tombs  "))+
    geom_col(data=aorist_KNC_pithoi,aes(decade,aoristic_density,fill=" KNC  "))+
      geom_col(data=aorist_FOR_pithoi,aes(decade,aoristic_density, fill=" Fortetsa  "))+
        theme_bw()+theme(legend.title=element_blank(),
          text=element_text(family="Garamond", size=12))+
            labs(x="Date (Years BC)", y="Aoristic Sum (Pithoi)")+
              scale_fill_brewer(palette = "Set2")+ 
                  scale_y_continuous(expand=c(0,0),limits=c(0,45))+
                    scale_x_reverse(expand = c(0, 0),limits=c(1100,600))+
                      theme(plot.margin=unit(c(0.1,1,0.1,0.1),"cm"))

aorist_pots_plot<-ggplot(NULL)+
  geom_col(data=aorist_all_tombs,aes(decade,aoristic_density,fill=" All tombs  "))+
    geom_col(data=aorist_KNC,aes(decade,aoristic_density,fill=" KNC  "))+
      geom_col(data=aorist_FOR,aes(decade,aoristic_density, fill=" Fortetsa  "))+
        theme_bw()+theme(legend.title=element_blank(),
          text=element_text(family="Garamond", size=12))+
            labs(x="Date (Years BC)", y="Aoristic Sum (All Vessels)")+
              scale_fill_brewer(palette = "Set2")+ 
                  scale_y_continuous(expand=c(0,0),limits=c(0,225))+
                    scale_x_reverse(expand = c(0, 0),limits=c(1100,600))+
                      theme(plot.margin=unit(c(0.1,1,0.1,0.1),"cm"))

ggarrange(aorist_pithoi_plot,aorist_pots_plot,ncol=2,common.legend=TRUE, legend="bottom")

@@

Figure \@ref(fig:aoristbycem) depicts the aoristic sums of pithoi (as a proxy for burials) and of all vessels, colour-coded by cemetery. Immediately apparent are the broad correspondences between both cemeteries. Despite differences in scale and timing, we can see the Fortetsa and Knossos North Cemeteries underwent similar chronological developments in the volume of burials and, to a lesser degree, grave goods. In both cemeteries, the rate of burial began rising around the end of the 9th century, plateaued slightly in the early 8th, and then rose again to a peak in the late 8th and early 7th centuries. For associated vessels, the developments are more volatile, but again there is clear growth beginning in the later 10th century - reaching an especially marked peak in the KNC by the end of the 9th - then a slight decline and levelling into the 8th century, before a final early 7th century apogee.

The near total absence of pithoi from the 11th and early 10th centuries reflects the delayed onset of cremation burial, with inhumations representing a significant proportion of earlier interments. Nonetheless, the clear inference would appear to be that, at both cemeteries, the late 8th and early 7th centuries witnessed an unprecedented level of burial activity, something which makes the sudden and total drop of chamber-tomb burials around 630 BC even more remarkable.

@@

```{r aorist_uncertainty, fig.dim = c(7, 4), dpi=300}

```{r aoristmedianadjust}

aorist_pot_medians$aorist_sums<-aorist_all_tombs$aoristic_density
aorist_pot_medians$aorist_median_adjust<-aorist_pot_medians$aorist_sums/aorist_pot_medians$aoristic_median
View(aorist_pot_medians)

ggplot(NULL)+geom_col(data=aorist_pot_medians,aes(decade,aorist_median_adjust))+
    xlim(1100,630)+theme_minimal()+
      theme(legend.title=element_blank(),
        text=element_text(family="Garamond", size=12))+
            labs(x="Date (Years BC)", y="Aoristic Median")+
               scale_fill_brewer(palette = "Set2")

```{r }

ggplot(aorist_melt,aes(variable,value, fill=period))+geom_boxplot()+theme_bw()+
      theme(legend.title=element_blank(),
        text=element_text(family="Garamond", size=12),
        axis.text.x = element_text(angle = 70, hjust = 1),
        panel.grid.major.x = element_blank())+
            labs(x="Date (Years BC)", y="Aoristic Sums per Tomb")+
               scale_fill_brewer(palette = "Set3")+
                scale_y_continuous(expand = c(0, 0),limits=c(0,20))

```{r }
tombs_gini<-matrix(0, ncol = 0, nrow = 1)
tombs_gini<-data.frame(tombs_gini)
for(i in colnames(aorist_tomb_values[2:51])) {
  tombs_gini[[paste0(i)]] <- ineq(aorist_tomb_values[[i]],type="Gini")
}

ggplot() + 
  geom_rect(data = rects, aes(xmin = xstart, xmax = xend, ymin = -Inf, ymax = Inf, fill = col), alpha = 0.4) +
     geom_line(data = tombs_gini, aes(decade,gini_coefficient),size=1)+
      theme_bw()+theme(legend.title=element_blank(),
          text=element_text(family="Garamond", size=12))+
              labs(x="Date (Years BC)", y="Gini Coefficient")+
                 scale_fill_brewer(palette = "Set3")+ coord_cartesian(
                    xlim = c(1100, 600), ylim = c(0.5,0.8))+ scale_x_reverse(expand = c(0, 0),limits=c(1100,600))

@@
